Interactive Visual Analysis of Image-Centric Cohort Study Data

### 2014-03-27 draft.tex
Incorporated Henrys feedback, two new figures

%% Abstract section.
\abstract{%
%(The problem). 
Epidemiological population studies impose information about a set of subjects (a \emph{cohort}) to characterize disease-specific risk factors.
%
Cohort studies comprise heterogenous variables (\emph{features}) describing the medical condition as well as demographic and lifestyle factors.
%
The data is analyzed using a priori defined hypotheses to find statistically significant correlations between variables (\emph{interactions}).
%
Modern cohort studies also incorporate medical image data.
%
%Analyzing these data requires image segmentation, extraction of key figures, and shape based subject grouping.
%
The statistically driven epidemiological workflow only allows to determine \emph{interactions} between image-derived metrics such as distances extracted from landmarks of the segmentation model.
\\\\
%(New Solution).
We propose an Interactive Visual Analysis (\emph{IVA}) approach that enables epidemiologists to examine both image-based as well as non-image data, e.g., sociodemographic variables and attributes derived from the image data.
%
This is achieved by combining brushing and linking enabled coordinated information visualization views and interactive 3D shape renderings with epidemiological data representations such as pivot tables and key figures as association measures.
%incorporating coordinated views from information visualization and 3D-shape renderings are combined with epidemiological data representations such as Pivot Tables.
%
%Multiple coordinated views from information visualization and 3D-shape renderings are combined with epidemiological data representations such as Pivot Tables.
%
%(Validation).
The presented concepts are applied expert-guided to gather and evaluate hypotheses about the aging process of the lumbar spine.
%
%(Results). 
It shows to be a more flexible comparison between image and non-image data.
%
%(Implications).
The new framework allows for hypothesis validation and hypothesis generation by incorporating human pattern recognition as well as data mining methods.
%
Using all reliable information from the image segmentation linked to non-image variables aims to unveil \emph{interactions} by applying an iterative analysis approach.

%
%Similarity measures between data variables are used to compute interesting changes in variable interactions for the current variable selection.
%
%Shape based grouping of subjects is facilitated using clustering techniques which operate on surface meshes extracted from the image segmentation.
} % end of abstract

% \abstract{Epidemiological population studies impose information about a set of subjects (a \emph{cohort}) to characterize disease specific risk factors.
% %
% Cohort studies comprise heterogenous variables (\emph{features}) describing the medical condition as well as demographic and lifestyle factors.
% %
% %Using well established statistical methods the data is hypothesis driven analyzed to find statistically significant variable correlations (\emph{interactions}).
% Driven by a priori defined hypotheses, the data is analyzed using well established statistical methods in order to find statistically significant correlations between variables (\emph{interactions}).
% %
% Modern cohort studies also incorporate medical image data.
% %
% Analyzing these data requires image segmentation, extraction of key figures, and shape based subject grouping.
% %
% The statistically driven epidemiological workflow only allows to determine \emph{interactions} between image-derived metrics such as distances extracted from landmarks of the segmentation model.
% \\\\
% We propose an Interactive Visual Analysis approach that enables epidemiologists to examine both image-based as well as non-image data, e.g., sociodemographic variables and attributes derived from the image data.
% %
% Using all reliable information from the image segmentation linked to non-image variables aims to unveil \emph{interactions} by applying an iterative analysis approach.
% %As extension to the epidemiological workflow it aims to unveil interactions based on all reliable shape information rather than abstracting it to a few key figures.
% %Iterative analysis between ... allows for extracting interactions which without the visiual analysis driven approach maybe would be overlooked. %Einsicht in Interaktionen zwischen nicht-bilddaten und bilddaten auf basis aller verfügbarer Informationen 
% %Iterative Analyse von Organstrukturen im Bezug auf Bilddaten 
% %Iterative analysis of image-based epidemiological datasets aims to provide insight into interactions between organ shape and % in order to provide insights into \emph{interactions} by incorporating additional information
% %It extends the epidemiological workflow with an iterative Analysis process and 
% It allows for hypothesis validation and hypothesis generation by incorporating human pattern recognition as well as data mining methods.
% %
% Multiple coordinated views from information visualization and 3D-shape renderings are combined with epidemiological data representations such as Pivot Tables.
% %
% Similarity measures between data variables are used to compute interesting changes in variable interactions for the current variable selection.
% %
% Shape based grouping of subjects is facilitated using clustering techniques which operate on surface meshes extracted from the image segmentation.
% } % end of abstract

@@

%% \section{Introduction} %for journal use above \firstsection{..} instead
Epidemiology aims at characterizing health and disease by determining risk factors.
%
Clinical problems, such as the selection of diagnostic tools and efficient treatment, are tackled using results of epidemiological research.
%
Also, the introduction of preventive measures in medicine and beyond, are based on epidemiological research, where, for example, subgroups with increased risk are identified \cite{Fletcher2012}.
%
On the other hand, observations made by clinicians in the daily routine are translated into hypothesis for epidemiological research.
%
These are used to determine environmental and lifestyle factors as well as medical attributes which may influence a condition of interest.
%
The data variables necessary are gathered using structured interviews and clinical examinations.
%
Statistical methods like regression analysis aim to check the attribute list for plausibility.
%
\\\\
Longitudinal population-based studies, such as the Study of Health in Pomerania (SHIP) \cite{Volzke2011}, aim to gather as much information as possible about a defined sample of people (a \emph{cohort}).
%
% The sample is drawn randomized to avoid selection \cite{Fletcher2012}.
% %
% When for example one investigates risk factors for prostate cancer in male subjects, the outcome is strongly dependent on the age.
% %
% Therefore results need to be age-adjusted to be comparable.
% %
% Confounding variables, are often not obvious at all and characterizing them is already an epidemiological result.
%
Modern cohort studies often include medical image data.
%
These data needs to be segmented to analyze its features for correlations with non-image data.
%
Manual segmentation is not feasible since it leads to a high variability which is at odds with the high reproducibility required in epidemiology research.
%
Semi-automatic techniques are more promising but challenging since the used methods as MRI and ultrasound are subject to inhomogeneity and noise.
%
Analyzing spatial data with respect to other epidemiological factors requires techniques which reach beyond standard statistical methods.

@@

	\item describing an Interactive Visual Analysis workflow for analyzing image-based epidemiological data including both hypothesis-driven and for hypothesis-generation based on a characterization of the standard epidemiological workflow
	\item providing visualization techniques which combine both information visualization and 3D rendering of organ shapes as well as combining them with well known epidemiological graphics and key figures.

@@

	\item implementing the presented methods as a web framework based on WebGL, D3JS and NodeJS.

@@

In this section we want to give insight into the epidemiological workflow when analyzing cohort study data to identify the problems we address in this paper.
%
% TODO Define Epidemiological Outcome - 5Ds:
% Death
% Disease
% Discomfort
% Disability
% Dissatisfaction
%
\subsection{Epidemiological Workflow} \label{EpidemiologicalWorkflow}
%Epidemiology diversity is reflected in the different experts working at cohort studies, ranging from specialized doctors to medical computer scientists with focus on biometrics and statisticians.
The diversity of epidemiology is reflected in the different experts who work at cohort studies, ranging from specialized doctors to medical computer scientists with focus on biometrics and statisticians.
%Many different experts work at epidemiological studies, ranging from specialized doctors to medical computer scientists with focus on biometrics and statisticians.
%
Epidemiologists follow a workflow mainly driven by statistic tools to validate hypothesis about disease specific risk factors.
%
Following Thew and colleagues, the workflow can be characterized as follows \cite{Thew2009}.
%
\begin{itemize}
	\item Hypotheses are based most commonly on observations made by clinicians in their daily routine.
%
	\item A set of attributes depicting conditions affected by the hypothesis is compiled accordingly.
% TODO: Bernhard: Effect size of Attribute ?
	\item Confounding features are adjusted so that they do not affect the effect size of a attribute.
%
	\item Statistical methods such as regression analysis are applied to measure the effect size of attributes to the outcome of interest.
\end{itemize}
The workflow is shown in Figure~\ref{fig:WorkflowComparison} (a).

@@

Grouping subjects using epidemiological features is essential in order to allow per-group risk determination.
%
%Grouping is carried out hypothesis driven.
Grouping depends on the underlying hypothesis.
%
Age for example is divided into groups (e.g. in 20 year-steps) when investigating its influence on a condition.
%
These groups depend strongly on the condition of interest and therefore there is no defined standard on how to categorize these values.

To determine, whether a subject is prone to be affected by a certain disease, \emph{relative risks} are expressed through the evaluation of p-values which indicate statistical significance.
%
Statistical correlations are prone to \emph{confounding}, meaning that the interaction of two features are influenced by a third feature which needs to isolated.
% Statistical correlations are prone to \emph{confounding}, meaning that two features are dependent and therefore should be normalized with respect to each other.
%
Statistics tools such as \texttt{SPSS}\footnote{Product of IBM; \url{www.ibm.com/software/de/analytics/spss/}} and \texttt{Stata}\footnote{Product of Stata Corp.; \url{http://www.stata.com/}} play a major role for analyzing epidemiological data.
%
Graphic data representation is largely used to present results rather than gaining insight.

@@

Indicators for medical conditions as well as questions about a subjects' lifestyle are also often \emph{dichotomous}--they have two manifestations (often \emph{Yes} or \emph{No}).
%
This allows for the calculation of \emph{odds ratios} which describe the relation of two \emph{dichotomous} variables, allowing for direct comparison of their influences.
%
Dichotomous data can also be derived by aggregating features to yield only two manifestations (e.g. subjects younger or older than 50 years).
%

\paragraph{Image acquisition.} \label{ImageAcquisition} Imaging techniques emitting hazardous amounts of radiation for the subject are not suited for ethical reasons.
%
MRI data is more expensive to obtain than CT data but does not affect the subjects health and is therefore the main method for collecting cohort study imaging data.
%
The image quality is a tradeoff between accuracy and affordability \cite{Preim2014}.
%
This often yields image resolutions inferior to those of clinical day-to-day practice, which makes their analysis more challenging.
%
The equipment used to gather medical image data is kept on the initial software and hardware version to ensure comparability in and between acquisition cycles.

@@

\subsection{The Study of Health in Pomerania (SHIP)}
After the pioneering Rotterdam study (started in 1990) several image-centric cohort study initiatives have evolved, they slightly differ in clinical focus, acquired data and epidemiologic research questions.
%
Starting 1997 with a cohort consisting of 4.308 subjects, the SHIP, located in northern Germany, aims to characterize health and disease in the widest range possible \cite{Volzke2011}.
%
Data is collected without focus on a group of diseases.
%
This allows the data set to be queried regarding many different diseases and conditions.
%
Subjects were examined in a 5-year time span, continuously adding new parameters including MRI scans in the last iteration of 2012.
%
The MRI protocol features a rich number of sequences.
%
%In addition to other MRI data, breast MRI scans were acquired for women.
%
A second cohort \texttt{SHIP-Trend} was established in 2008 to acquire data about a younger population.
%
The protocols for analyzing the subjects between \texttt{SHIP} and \texttt{SHIP-Trend} remained the same, making them comparable.
%
The overall examination time for each person attending the study is two days.

@@

%\emph{Over-fitting} of hypotheses to the data may be imposed  is a danger pointed out by the authors when hypotheses to the data based on observations of changes in these plots the measure highlights only subsets of statistical changes.
Hypotheses based on observations of changes in these plots may impose \emph{overfitting} to the data because the measure highlights only subsets of statistical changes.
%
Our approach sticks more to the information extracted from the segmented image data and derive variable interaction with non-image epidemiological factors.
% Disadvantage - Nur 2 abgeleitete Variablen sichtbar, schlecht mit anderen Variablen in Verbindung zu bekommen (müsste jedoch auch gehen) - viele Dinge gehen durch die Abstraktion aber auch flöten!

@@

\paragraph{Visualizing Heterogenous Non-Image Data.}
Generalized Pairs Plots (\texttt{GPLOM´S}) are an information visualization technique comparing heterogenous variables pairwise using a plot-matrix grouped by type \cite{Francois2013}.
%
It is useful to gain an overview over numerous variables and their distributions.
%This technique is also useful to gain an overview over numerous variables and their distributions.
%
Histograms, bar charts, scatterplots and heat maps are used to visualize variable combinations with regard to their type.
%
%Brushing, linking and filtering are feasible with \texttt{GPLOMS}, but have limitations such as making only one category brushable at a time.
%
%We applied this technique to our data and it shows promising potential for simultaneously visualizing many different variables. %but does not fit in the scope of this paper.
%
%The inspiration on the chosen visualization techniques stems from this publication.
%
Dai and colleagues explored risk factors by incorporating choropleth maps of epidemiological features (e.g. mortality rates in a region) with parallel coordinates, bar charts and scatterplots with integrated regression lines \cite{Dai2005}.
%A similar approach was taken by Dai and colleagues for risk factor exploration as they also incorporate choropleth maps of epidemiological factors (e.g. mortality rates in a region) with parallel coordinates, bar charts and scatterplots with integrated regression lines \cite{Dai2005}.
%
Their findings yielded a \emph{Concept Map} which linked cancer-related interactions via graph edges.
%From their findings regarding the interaction of cancer-related socio-demographic factors are drawn in a \emph{Concept Map} where related factors are connected via graph-edges.
Chui and colleagues visualized interactions in time-dependent epidemiological data using time-series plots highlighting risk factors differences in age and gender \cite{Chui2011}.

@@

%\section{Interactive Visual Analyis in Cohort Study Data}
\section{Image Centric Cohort Study Data in Interactive Visual Analysis Context} \label{Image Centric Cohort Study Data in Interactive Visual Analysis Context}
\begin{figure}[htb]
 \centering
 \includegraphics[width=3.0in]{figures/workflow_comparison}
 \caption{\emph{IVA} tools are able to complement parts of the epidemiological workflow, not replace it. The appropriate combination of statistical- and interactive driven analysis shows promising potential to unveil the information in the data. (a) shows the standard epidemiological workflow, (b) the IVA supported one. The iterative red highlighted part is called the \emph{IVA Loop}.}
  \label{fig:WorkflowComparison}
\end{figure}
%- Acceptance plays a big role - you have to pick up the user by presenting information they are used to in a new way\\
%
Subsection~\ref{EpidemiologicalWorkflow} described the epidemiological workflow as a sequence of steps taken by domain experts which need to comprise reproducibility and statistical integrity.
%
Figure~\ref{fig:WorkflowComparison} (a) shows this workflow as consecutive series of steps.
%
Introducing the \emph{IVA} principle to the epidemiological application domain does aim to complement its weaknesses rather than replacing the existing workflow.
%The workflow we propose by introducing the \emph{IVA} principle into the epidemiological application domain does not aim to replace the existing workflow but to complement its weaknesses.
%
In the current state the workflow treats the data like a black box.
%
Statistical tests on a hypothesis-describing feature list yields a value deciding whether the data supports the hypothesis or not.
%A list of features describing the hypothesis is compiled and analyzed using statistical tests. 
%
%The resulting value decides whether the data supports the hypothesis or not.
%
Features not included in the analysis may potentially support the chosen hypothesis by discriminating the population in the expected way, but are not highlighted in any way.
%It would be possible that there are actually features of the data set which support the hypothesis by discriminating the population in the expected way, but with this approach they are not highlighted in any way.
%
%Adapting this workflow to medical image data analysis also highlights this problem.
This becomes even more important when the workflow is adapted to the analysis of the medical image data, where
domain experts annotate landmarks tediously to derive metrics such as diameters. %which are then handled like other features and analyzed using the same set of statistical tools.
%
This leaves out the majority of information in the image data by abstracting it to single values.
%
It is easily possible that information left out would heavily influence the result.
%Not only does this leave out the majority of the information in the medical image data by abstracting it to single values, it is easily possible that information left out would heavily influence the result.
%
Considering more complex parts of the data would make those results more trustworthy and also could identify possible anatomical confounders--an epidemiolgical research result in itself.

Statistical tests check for validity of the number but not for their completeness or plausibility!
\\\\
\emph{IVA} tries to illuminate the black box by making the domain experts part of an iterative feature list selection process, which is shown in Figure~\ref{fig:WorkflowComparison} (b) as part of the epidemiological workflow.
%
%Figure~\ref{fig:WorkflowComparison} (b) highlights the iterative process as part of the epidemiolgical workflow.
%
It also aims to project back into the hypothesis formulation step to amplify hypothesis generation.
%Note that it also aims to project back into the hypothesis formulation step to amplify hypothesis generation 
%
This has to be handled with care since \emph{overfitting} of expectations to the data is an imminent danger as described by Turkay and colleagues \cite{Turkay2013}.

@@

\subsection{IVA Patterns}
%
\emph{IVA} knows three different exploration patterns, handling interaction of domain and range variables.
%The explorative procedures when analyzing data using \emph{IVA} can be divided into three different patterns, handling interaction between domain and range variables.

\paragraph{Local Investigation}
This pattern projects information from domain space to the range perspective.
%
This step is more complicated in epidemiological context compared to other \emph{IVA} application domains.
%As opposing to other \emph{IVA} application domains, this step is more complicated in the epidemiological context.
%
Shape information can not be brushed using ROI-selections but rather has to employ techniques that specify local deformation changes \cite{Hermann2014} or subjects that belong to a shape class.
%
\emph{Feature selection} methods strongly depend on the type of segmentation used to extract the tissue of interest.
%
Model-based segmentations or masks yield data structures capable of calculating mean shapes and distances between individuals or subject groups.
%
Feature selection is also possible by applying clustering algorithms in order to get shape-groups \cite{Klemm2013VMV}.
%
These algorithms can be used to investigate interactions between shape-groups and other non-image based variables.
%
Analysis of outliers can indicate segmentation errors or an group of subjects sharing a pathology.
%
%Another application is the outlier analysis. 
%
%Outliers can indicate segmentation errors or an outstanding group of individuals who may share a pathology.

@@

>>>>

%% Abstract section.
\abstract{%
%(The problem).
Epidemiological population studies impose information about a set of subjects (a \emph{cohort}) to characterize disease-specific risk factors.
%
Cohort studies comprise heterogenous variables (\emph{features}) describing the medical condition as well as demographic and lifestyle factors.
%
The data is analyzed using a priori defined hypotheses to find statistically significant correlations between variables (\emph{associations}).
%
Modern cohort studies also incorporate medical image data.
%
%Analyzing these data requires image segmentation, extraction of key figures, and shape based subject grouping.
%
The statistically driven epidemiological workflow only allows to determine \emph{associations} between image-derived metrics such as distances extracted from landmarks of the segmentation model.
\\\\
%(New Solution).
We propose an Interactive Visual Analysis (\emph{IVA}) approach that enables epidemiologists to examine both image-based as well as non-image data, e.g., sociodemographic variables and attributes derived from the image data.
%
This is achieved by combining brushing and linking enabled coordinated information visualization views and interactive 3D shape renderings with epidemiological data representations such as pivot tables and key figures as association measures.
%incorporating coordinated views from information visualization and 3D-shape renderings are combined with epidemiological data representations such as Pivot Tables.
%
%Multiple coordinated views from information visualization and 3D-shape renderings are combined with epidemiological data representations such as Pivot Tables.
%
%(Validation).
The presented concepts are applied expert-guided to gather and evaluate hypotheses about the aging process of the lumbar spine.
%
%(Results). 
It shows to be a more flexible comparison between image and non-image data.
%
%(Implications).
The new framework allows for hypothesis validation and hypothesis generation by incorporating human pattern recognition as well as data mining methods.
%
Using all reliable information from the image segmentation linked to non-image variables aims to unveil \emph{associations} by applying an iterative analysis approach.

%
%Similarity measures between data variables are used to compute interesting changes in variable associations for the current variable selection.
%
%Shape based grouping of subjects is facilitated using clustering techniques which operate on surface meshes extracted from the image segmentation.
} % end of abstract

% \abstract{Epidemiological population studies impose information about a set of subjects (a \emph{cohort}) to characterize disease specific risk factors.
% %
% Cohort studies comprise heterogenous variables (\emph{features}) describing the medical condition as well as demographic and lifestyle factors.
% %
% %Using well established statistical methods the data is hypothesis driven analyzed to find statistically significant variable correlations (\emph{associations}).
% Driven by a priori defined hypotheses, the data is analyzed using well established statistical methods in order to find statistically significant correlations between variables (\emph{associations}).
% %
% Modern cohort studies also incorporate medical image data.
% %
% Analyzing these data requires image segmentation, extraction of key figures, and shape based subject grouping.
% %
% The statistically driven epidemiological workflow only allows to determine \emph{associations} between image-derived metrics such as distances extracted from landmarks of the segmentation model.
% \\\\
% We propose an Interactive Visual Analysis approach that enables epidemiologists to examine both image-based as well as non-image data, e.g., sociodemographic variables and attributes derived from the image data.
% %
% Using all reliable information from the image segmentation linked to non-image variables aims to unveil \emph{associations} by applying an iterative analysis approach.
% %As extension to the epidemiological workflow it aims to unveil associations based on all reliable shape information rather than abstracting it to a few key figures.
% %Iterative analysis between ... allows for extracting associations which without the visiual analysis driven approach maybe would be overlooked. %Einsicht in Interaktionen zwischen nicht-bilddaten und bilddaten auf basis aller verfügbarer Informationen 
% %Iterative Analyse von Organstrukturen im Bezug auf Bilddaten 
% %Iterative analysis of image-based epidemiological datasets aims to provide insight into associations between organ shape and % in order to provide insights into \emph{associations} by incorporating additional information
% %It extends the epidemiological workflow with an iterative Analysis process and 
% It allows for hypothesis validation and hypothesis generation by incorporating human pattern recognition as well as data mining methods.
% %
% Multiple coordinated views from information visualization and 3D-shape renderings are combined with epidemiological data representations such as Pivot Tables.
% %
% Similarity measures between data variables are used to compute interesting changes in variable associations for the current variable selection.
% %
% Shape based grouping of subjects is facilitated using clustering techniques which operate on surface meshes extracted from the image segmentation.
% } % end of abstract

@@

%% \section{Introduction} %for journal use above \firstsection{..} instead
Epidemiology aims at characterizing health and disease by determining risk factors.
%
Clinical problems, such as the selection of diagnostic tools and efficient treatment, are tackled using results of epidemiological research.
%
Also, the introduction of preventive measures in medicine and beyond, are based on epidemiological research, where, for example, subgroups with increased risk are identified \cite{Fletcher2012}.
%
On the other hand, observations made by clinicians in the daily routine are translated into hypothesis for epidemiological research.
%
These are used to determine environmental and lifestyle factors as well as medical attributes which may influence a condition of interest.
%
The data variables necessary are gathered using structured interviews and clinical examinations.
%
Methods like regression analysis check the attribute list for statistical soundness.
%
\\\\
Longitudinal population-based studies, such as the Study of Health in Pomerania (SHIP) \cite{Volzke2011}, gather as much information as possible about a defined sample of people (a \emph{cohort}).
%
% The sample is drawn randomized to avoid selection \cite{Fletcher2012}.
% %
% When for example one investigates risk factors for prostate cancer in male subjects, the outcome is strongly dependent on the age.
% %
% Therefore results need to be age-adjusted to be comparable.
% %
% Confounding variables, are often not obvious at all and characterizing them is already an epidemiological result.
%
Cohort studies often include medical image data.
%
These data needs to be segmented to analyze its features for correlations with non-image data.
%
Manual segmentation is not feasible since it leads to a high variability which is at odds with the high reproducibility required in epidemiology research.
%
Semi-automatic techniques are more promising but challenging since the used methods as MRI and ultrasound are subject to inhomogeneity and noise.
%
Analyzing spatial data with respect to other epidemiological factors requires techniques which reach beyond standard statistical methods.

@@

	\item an Interactive Visual Analysis workflow for analyzing image-based epidemiological data including both hypothesis-driven and for hypothesis-generation based on a characterization of the standard epidemiological workflow
	\item visualization techniques, which combine both information visualization and 3D rendering of organ shapes as well as combining them with epidemiological graphics and key figures.

@@

	\item implementing the presented methods as a web framework based on \texttt{WebGL}, \texttt{D3.js} and \texttt{NodeJS}.

@@

In this section we give insight into the epidemiological workflow when analyzing cohort study data to identify the problems we address in this paper.
%
% TODO Define Epidemiological Outcome - 5Ds:
% Death
% Disease
% Discomfort
% Disability
% Dissatisfaction
%
\subsection{Epidemiological Workflow} \label{EpidemiologicalWorkflow}
%Epidemiology diversity is reflected in the different experts working at cohort studies, ranging from specialized doctors to medical computer scientists with focus on biometrics and statisticians.
The diversity of epidemiology is reflected in the different experts who work at cohort studies, ranging from specialized doctors to medical computer scientists with focus on biometrics and statisticians.
%Many different experts work at epidemiological studies, ranging from specialized doctors to medical computer scientists with focus on biometrics and statisticians.
%
Epidemiologists follow a workflow mainly driven by statistic tools to validate hypothesis about disease specific risk factors.
%
Following Thew and colleagues, the workflow can be characterized as follows \cite{Thew2009}.
%
\begin{itemize}
	\item Hypotheses are based most commonly on observations made by clinicians in their daily routine.
%
	\item A set of attributes depicting conditions affected by the hypothesis is compiled accordingly.
% TODO: Bernhard: Effect size of Attribute ?
	\item Confounding features are adjusted so that they do not affect the effect size of a attribute.
%
	\item Statistical methods such as regression analysis are applied to measure the effect size of attributes to the outcome of interest.
\end{itemize}
The workflow is shown in Figure~\ref{fig:WorkflowComparison} (a).

@@

In longitudinal cohort design, grouping subjects using epidemiological features is essential in order to allow per-group risk determination.
%
%Grouping is carried out hypothesis driven.
Grouping depends on the underlying hypothesis.
%
Age for example is divided into groups (e.g. in 20 year-steps) when investigating its influence on a condition.
%
These groups depend strongly on the condition of interest and therefore there is no standard on how to categorize these.

To determine, whether a subject is prone to be affected by a certain disease, \emph{relative risks} are expressed through the evaluation of p-values which indicate statistical significance.
%
Statistical correlations are prone to \emph{confounding}, meaning that the association of two features are influenced by a third feature which needs to isolated.
% Statistical correlations are prone to \emph{confounding}, meaning that two features are dependent and therefore should be normalized with respect to each other.
%
%Statistics tools such as \texttt{SPSS}\footnote{Product of IBM; \url{www.ibm.com/software/de/analytics/spss/}} and \texttt{Stata}\footnote{Product of Stata Corp.; \url{http://www.stata.com/}} play a major role for analyzing epidemiological data.
Statistics tools such as \texttt{SPSS}\footnote{Product of IBM; \url{www.ibm.com/software/de/analytics/spss/}} play a major role for analyzing epidemiological data.
%
Graphic data representation is largely used to present results rather than gaining insight.

@@

Indicators for medical conditions as well as questions about a subjects' lifestyle are also often \emph{dichotomous}--they have two manifestations (\emph{Yes} or \emph{No}).
%
%This allows for the calculation of \emph{odds ratios} which describe the relation of two \emph{dichotomous} variables, allowing for direct comparison of their influences.
%
Dichotomous data can also be derived by aggregating features to yield only two manifestations (e.g. subjects younger or older than 50 years).
%

\paragraph{Image acquisition.} \label{ImageAcquisition} Imaging techniques emitting hazardous amounts of radiation for the subject are not suited for ethical reasons.
%
%MRI data is more expensive to obtain than CT data but does not affect the subjects health and is therefore the main method for collecting cohort study imaging data.
MRI the main method for collecting cohort study imaging data.
%
The image quality is a tradeoff between accuracy and affordability \cite{Preim2014}.
%
This often yields image resolutions inferior to those of clinical day-to-day practice, which makes their analysis more challenging.
%
The equipment used to gather medical image data is kept, if possible, on the initial software and hardware version to ensure comparability in and between acquisition cycles.

@@

\subsection{The Study of Health in Pomerania (SHIP)}
After the pioneering Rotterdam study (started in 1990) several MR imaging studies initiatives have evolved, they slightly differ in clinical focus, acquired data and epidemiologic research questions.
%
Starting 1997 with a cohort consisting of 4.308 subjects, the SHIP, located in northern Germany, aims to characterize health and disease in the widest range possible \cite{Volzke2011}.
%
Data is collected without focus on a group of diseases.
%
This allows the data set to be queried regarding many different diseases and conditions.
%
Subjects were examined in a 5-year time span, continuously adding new parameters including MRI scans in the last iteration of 2012.
%
The MRI protocol features a rich number of sequences.
%
%In addition to other MRI data, breast MRI scans were acquired for women.
%
%A second cohort \texttt{SHIP-Trend} was established in 2008 to acquire data about a younger population.
A second cohort \texttt{SHIP-Trend} was established in 2008.
%
The protocols for analyzing the subjects between \texttt{SHIP} and \texttt{SHIP-Trend} remained the same, making them comparable.
%
The overall examination time for each person attending the study is two days.

@@

%\emph{Over-fitting} of hypotheses to the data may be imposed  is a danger pointed out by the authors when hypotheses to the data based on observations of changes in these plots the measure highlights only subsets of statistical changes.
Hypotheses based on observations of changes in these plots may impose \emph{overfitting} to the data because the measure highlights only subsets of statistical changes.
%
Our approach sticks more to the information extracted from the segmented image data and derive variable associations with non-image epidemiological factors.
% Disadvantage - Nur 2 abgeleitete Variablen sichtbar, schlecht mit anderen Variablen in Verbindung zu bekommen (müsste jedoch auch gehen) - viele Dinge gehen durch die Abstraktion aber auch flöten!

@@

\paragraph{Visualizing Heterogenous Non-Image Data.}
Generalized Pairs Plots (\texttt{GPLOM´S}) are an information visualization technique comparing heterogenous variables pairwise using a plot-matrix grouped by type \cite{Francois2013}.
%
It is useful to gain an overview over numerous variables and their distributions.
%This technique is also useful to gain an overview over numerous variables and their distributions.
%
Histograms, bar charts, scatterplots and heat maps are used to visualize variable combinations with regard to their type.
%
%Brushing, linking and filtering are feasible with \texttt{GPLOMS}, but have limitations such as making only one category brushable at a time.
%
%We applied this technique to our data and it shows promising potential for simultaneously visualizing many different variables. %but does not fit in the scope of this paper.
%
%The inspiration on the chosen visualization techniques stems from this publication.
%
Dai and colleagues explored risk factors by incorporating choropleth maps of epidemiological features (e.g. mortality rates in a region) with parallel coordinates, bar charts and scatterplots with integrated regression lines \cite{Dai2005}.
%A similar approach was taken by Dai and colleagues for risk factor exploration as they also incorporate choropleth maps of epidemiological factors (e.g. mortality rates in a region) with parallel coordinates, bar charts and scatterplots with integrated regression lines \cite{Dai2005}.
%
Their findings yielded a \emph{Concept Map} which linked cancer-related associations via graph edges.
%From their findings regarding theassociationn of cancer-related socio-demographic factors are drawn in a \emph{Concept Map} where related factors are connected via graph-edges.
Chui and colleagues visualized associations in time-dependent epidemiological data using time-series plots highlighting risk factors differences in age and gender \cite{Chui2011}.

@@

%\section{Interactive Visual Analyis in Cohort Study Data}
\section{Image Centric Cohort Study Data in Interactive Visual Analysis Context} \label{Image Centric Cohort Study Data in Interactive Visual Analysis Context}
\begin{figure}[htb]
 \centering
 \includegraphics[width=3.0in]{figures/workflow_comparison}
 \caption{\emph{IVA} tools are able to complement parts of the epidemiological workflow, not replace it. The appropriate combination of statistical- and interactive driven analysis shows promising potential to unveil the information in the data. (a) shows the standard epidemiological workflow, (b) the IVA supported one. The iterative red highlighted part is called the \emph{IVA Loop}.}
  \label{fig:WorkflowComparison}
\end{figure}
%- Acceptance plays a big role - you have to pick up the user by presenting information they are used to in a new way\\
Subsection~\ref{EpidemiologicalWorkflow} described the epidemiological workflow as a sequence of steps taken by domain experts which need to comprise reproducibility and statistical integrity.
Figure~\ref{fig:WorkflowComparison} (a) shows this workflow as consecutive series of steps.
Introducing the \emph{IVA} principle to the epidemiological application domain does aim to complement its weaknesses rather than replacing the existing workflow.
%The workflow we propose by introducing the \emph{IVA} principle into the epidemiological application domain does not aim to replace the existing workflow but to complement its weaknesses.
In the current state the workflow treats the data like a black box.
Statistical tests on a hypothesis-describing feature list yields a value deciding whether the data supports the hypothesis or not.
%A list of features describing the hypothesis is compiled and analyzed using statistical tests. 
%The resulting value decides whether the data supports the hypothesis or not.
Features not included in the analysis may potentially support the chosen hypothesis by discriminating the population in the expected way, but are not highlighted in any way.
%It would be possible that there are actually features of the data set which support the hypothesis by discriminating the population in the expected way, but with this approach they are not highlighted in any way.
%Adapting this workflow to medical image data analysis also highlights this problem.
This becomes even more important when the workflow is adapted to the analysis of the medical image data, where
domain experts annotate landmarks tediously to derive metrics such as diameters. %which are then handled like other features and analyzed using the same set of statistical tools.
This leaves out the majority of information in the image data by abstracting it to single values.
It is easily possible that information left out would heavily influence the result.
%Not only does this leave out the majority of the information in the medical image data by abstracting it to single values, it is easily possible that information left out would heavily influence the result.
Considering more complex parts of the data would make those results more trustworthy and also could identify possible anatomical confounders--an epidemiolgical research result in itself.
%Statistical tests check for validity of the number but not for their completeness or plausibility!
\emph{IVA} tries to illuminate the black box by making the domain experts part of an iterative feature list selection process, which is shown in Figure~\ref{fig:WorkflowComparison} (b) as part of the epidemiological workflow.
%Figure~\ref{fig:WorkflowComparison} (b) highlights the iterative process as part of the epidemiolgical workflow.
It also aims to project back into the hypothesis formulation step to amplify hypothesis generation.
%Note that it also aims to project back into the hypothesis formulation step to amplify hypothesis generation 
This has to be handled with care since \emph{overfitting} of expectations to the data is an imminent danger as described by Turkay and colleagues \cite{Turkay2013}.

@@

\subsection{IVA Patterns}
%
\emph{IVA} knows three different exploration patterns, handling associations of domain and range variables.
%The explorative procedures when analyzing data using \emph{IVA} can be divided into three different patterns, handlinassociationon between domain and range variables.

\paragraph{Local Investigation}
This pattern projects information from domain space to the range perspective.
%
This step is more complicated in epidemiological context compared to other \emph{IVA} application domains.
%As opposing to other \emph{IVA} application domains, this step is more complicated in the epidemiological context.
%
Shape information can not be brushed using ROI-selections but rather has to employ techniques that specify local deformation changes \cite{Hermann2014} or subjects that belong to a shape class.
%
\emph{Feature selection} methods strongly depend on the type of segmentation used to extract the tissue of interest.
%
Model-based segmentations or masks yield data structures capable of calculating mean shapes and distances between individuals or subject groups.
%
Feature selection is also possible by applying clustering algorithms in order to get shape-groups \cite{Klemm2013VMV}.
%
These algorithms can be used to investigate associations between shape-groups and other non-image based variables.
%
Analysis of outliers can indicate segmentation errors or an group of subjects sharing a pathology.
%
%Another application is the outlier analysis. 
%
%Outliers can indicate segmentation errors or an outstanding group of individuals who may share a pathology.

@@

\begin{figure}[htb]
 \centering
 \includegraphics[width=3.5in]{figures/similarity_matrix}
 \caption{Adjacency matrix of 131 features with a grand total of 17.161 combinations. Similarity is calculated using the \emph{Cram\'{e}rs' V} contingency value. Color brightness encodes association strength. The enlarged excerpt shows associations for shape clusters of subjects with and without diagnosed spine attrition, which show associations between gender, weight, body height and smoking behavior.
 }
 \label{fig:similarity}
\end{figure}
